An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2015-04-20 Response.txt

>>> 2. The explanation of the inter-model proposal and acceptance criterion is Section 3c is
inadequate. While well-known material on Bayesian analysis (Section 3a) and MCMC
(Section 3b) is presented at length with equations, the new RJMCMC approach is
skimmed over in a single sentence (no equations) as \We perform MCMC in the supermodel
parameter space just like a regular MCMC; we propose jumps to different parameters
within a model (intra-model jumps) and jumps to a different model with
different parameters (inter-model jumps)" (page 6)...

We have added descriptive text at the bottom of page 5 and top of page 6 that expands on the RJMCMC approach.  The text now makes clear that the jump proposal ratio is a ratio of densities in the two model spaces, and discusses the special case where the spaces are of equal dimension and the proposal is a diffeomorphism between them (i.e.\ not a probabilistic choice of jump point, but rather a fixed mapping), where the ratio reduces to the Jacobian factor to which the referee refers.  We have not written the equation as given by the referee,

p_accept = (p(theta')*Q(theta' -> theta))/(p(theta)*Q(theta -> theta')) * |J|

because this form is not correct in our notation.  The factors Q *already incorporate the Jacobian* in our notation.  This point is now emphasised in the text.

@@

>>> 3. The examples given in the manuscript seem simplified to the point of triviality and/or
are poorly explained. The only example given of model selection (the potentially important
contribution of this paper) only involves sampling between 1D Gaussian and
Cauchy distributions (two unknown parameters each). While this does illustrate the
method (barely), it is of no practical relevance and is so much simpler than most realistic
model-selection problems that it is not clear how efficiency and performance
would scale, or how useful the method could be in practice. I assume the authors are
researching real-world model-selection problems of practical interest (like alternative
mass distribution models for black-hole X-ray binaries, mentioned on page 4). Including
a meaningful example such as this would greatly improve the paper and increase
its significance. The other two examples represent only single-model sampling and are
of much less interest, but even so they should be properly explained. In the examples
in Figures 3 and 4, what are the problems, what are the parameters, and what do the
results mean? This current presentation is not satisfying to the reader.

We have added text to the beginning of Section 4 pointing to the use of this algorithm in real-world applications, such as those of Farr et al (2011).  We feel that to describe the (lengthy) analysis from that work would needlessly complicate the presentation of the algorithm in this paper, but hope that this emphasises that the algorithm is, indeed, applicable to complicated model selection problems.  We also emphasise in that text that the algorithm is available as a software library for others interested in adopting it.

We have added text to Figures 3 and 4 describing the gravitational wave data analysis problem and the associated parameter space briefly.  Again, we feel that a complete description of this problem (available from Veitch et al (2015) and the other references in the text) is beyond the scope of the article, but we understand the need for more context in the explanation of this paper.  We hope the additional text provides this context.

@@

>>> 4. Claims are made of \dramatically improved convergence" (page 3), that \convergence
is generally rapid" (page 4), and \the interpolation method described above can dramatically
improve the runtime of an RJMCMC" (page 10). While I believe the method
developed here could substantially improve efficiency in some cases, this is not actually
demonstrated in the paper. The trivial example in Figure 2 appears to increase
inter-model acceptance by a factor of 8 or so, but no results on RJMCMC sampling
convergence are given. The example considered in Figure 4 claims to have improved
convergence by a factor of two (not dramatic), but there is no discussion of what definition is applied for convergence|and as noted on page 6 (third paragraph), defining
convergence for MCMC sampling is a tricky business. Even applying a reasonable
working definition of convergence, given the random nature of MCMC sampling, averages
over several MCMC runs are often required to meaningfully compare convergence
rates -- was that done here? While the claims of \dramatic" improvements in convergence may be true in some cases, they don't appear to be properly established in the paper.

We have removed the (perhaps over-wrought) adjective "dramatic" from our claims of convergence, and included text briefly describing the method of computing convergence times used and referring the reader for details to Veitch et al (2015).

### 2015-04-20 paper.tex
Changes to address first reviewer's comments + response.

\begin{abstract}
Selection among alternative theoretical models given an observed data set is an important challenge in many areas of physics and astronomy.  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an extremely powerful technique for performing Bayesian model selection, but it suffers from a fundamental difficulty: it requires jumps between model parameter spaces, but cannot efficiently explore both parameter spaces at once.  Thus, a naive jump between parameter spaces is unlikely to be accepted in the MCMC algorithm and convergence is correspondingly slow.  Here we demonstrate an interpolation technique that uses samples from single-model MCMCs to propose inter-model jumps from an approximation to the single-model posterior of the target parameter space.  The interpolation technique, based on a kD-tree data structure, is adaptive and efficient in modest dimensionality.  We show that our technique leads to dramatically improved convergence over naive jumps in an RJMCMC, and compare it to other proposals in the literature to improve the convergence of RJMCMCs.  We also demonstrate the use of the same interpolation technique as a way to construct efficient ``global'' proposal distributions for single-model MCMCs without prior knowledge of the structure of the posterior distribution, and discuss improvements that permit the method to be used in higher-dimensional spaces efficiently.
\end{abstract}

@@

Here we introduce an alternative technique based on a kD-tree data 
structure to construct an approximation
to each model's posterior parameter distribution.  This improved interpolation method leads to dramatically faster convergence of RJMCMC sample chains.
We draw jump proposals into the model from this approximation to its posterior.
Because jumps are proposed preferentially to locations favored by the
single-model posterior, the RJMCMC compares ``good'' locations in
parameter space across all the models, and convergence is generally
rapid.  We have successfully applied this RJMCMC technique to a 10-way
model selection among alternative mass distribution models for
black-hole X-ray binaries \cite{Farr2010}.  We also provide an example
using this method as an ``asymptotically Markovian''
\cite{terBraak2008} jump proposal in the context of a single-model,
nine-dimensional MCMC in Section \ref{sec:higherdim}.

@@

The parameter space of the super-model in an RJMCMC consists of a
discrete parameter that identifies the model, $M_i$, and a set of
continuous parameters appropriate for that model, $\vtheta_i$.  Thus,
each sample consists of a model identifier and a location within the
parameter space of that model, $\{M_i, \vtheta_i\}$.  We perform the
MCMC in the ``super-model'' parameter space just like a regular MCMC;
we propose jumps to different parameters within a model (intramodel
jumps) and jumps to a different model with different parameters
(intermodel jumps).  The resulting chain samples from the posterior
$p(M_i, \{\vtheta_i\}|d)$.  As in a usual MCMC, the PDF on the model as a
parameter, with other parameters ignored, is obtained by marginalizing
over the remaining parameters.  The posterior probability of a model
is proportional to the number of counts
%
\be
%
p(M_i|d) = \int d\vtheta_i \frac{L(d|M_i, \vtheta_i) p(\vtheta_i|M_i) p(M_i)}{p(d)}
\approx \frac{N_i}{N},
%
\ee
%
where $N_i$ is the number of RJMCMC samples listing the $i$'th model
and $N$ is the total chain length.  Thus, the probability of a
particular model relative to other models under consideration is given
by the fraction of RJMCMC samples lying in the parameter space of that
model.

@@

In this section, we demonstrate the efficiency of the kD-interpolated
jump proposal on a toy model-comparison problem.  We draw $N = 100$
simulated data points from a $N(0,1)$ Gaussian distribution, and then
ask whether these data are better described by a model where they are
Gaussian distributed with unknown mean $\mu$ and standard deviation
$\sigma$
%
\be
p(x) = \frac{1}{\sqrt{2\pi} \sigma} \exp\left( - \frac{(x-\mu)^2}{2
    \sigma^2} \right)\ ,
\ee
%
or by a model where they are Cauchy distributed with mode $\alpha$
and width $\beta$
%
\be
p(x) = \frac{1}{\pi \beta \left( 1 + \left(\frac{x - \alpha}{\beta}\right)^2\right)}\ .
\ee
%
We take priors on $\mu$ and $\alpha$ to be uniform in $[-1,1]$, and
priors in $\sigma$ and $\beta$ to be uniform in $[0.5, 1.5]$.  With a
data set of 100 points, the relative uncertainty in determining the
parameters of the underlying distribution is approximately 10\%, so we
expect the posterior probabilities in the $(\mu,\sigma)$ and
$(\alpha,\beta)$ spaces to occupy only a few percent of the prior
volume.  The Cauchy distribution is much broader than the Gaussian (it
has no finite moments), so with equal model priors, the posterior
probability for the Gaussian model over the Cauchy model is extremely
large:
%
\be
\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} \sim 10^9.
\ee
%
In order to ensure that the RJMCMC produces samples in the Cauchy
model at all, we impose a model prior that favors the Cauchy model by
$5 \times 10^8$ relative to the Gaussian.  The evidence ratio between
the models for our chosen data set with these priors is
%
\be \frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)}
\equiv r = 1.15,
\ee
%
yielding a theoretical maximum acceptance rate of inter-model jumps of
$(1+1/r)/2 = 0.93$.

@@

The relative error on the determination of the Bayes factor (evidence ratio) scales
with $1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of inter-model
transitions in the RJMCMC.  Thus, as the acceptance rate of
inter-model jumps goes down, the RJMCMC must run longer to achieve a
desired accuracy in the evidence ratio.  By boosting the acceptance
rate of inter-model jumps, the interpolation method described above
can dramatically improve the runtime of an RJMCMC.

@@

\begin{figure}
  \includegraphics[width=0.8\columnwidth]{PCC}
  \caption{\label{fig:PCC} A two-dimensional illustration of the two
    boxes involved in the modified kD proposal.  The larger box
    aligned with the axes is the normal kD-tree box containing the
    given samples, while the tighter box is the ``covariance cell'' aligned with the eigenvectors of the sample distribution.  The modified kd-interpolated jump proposal draws a point from the intersection of the two boxes. }
\end{figure}

@@

Figure \ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
proposal applied as part of a \texttt{LALInferenceMCMC} analysis.  In this example, the kD-tree jump proposal is one of several jump proposals used during the MCMC, and the kD tree itself is updated with accepted samples whenever its jump proposal is called. In spite of the very low acceptance
rate of the proposal, applying the kD-tree proposal to one out of
20 proposed jumps improved the convergence time of the simulation
by a factor of two compared to the standard suite of proposals because
it is particularly efficient at producing ``mode-hopping'' jumps,
which are difficult to produce with the other proposals in
\texttt{LALInferenceMCMC}.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative parameter acceptance
    ratio for the modified kD-tree jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). In spite of the small acceptance ratio, the
    kD tree proposal, applied to one in 20 proposed jumps,
    improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals. 
    The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated.
%\dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).  Is this a fluke, or do you always see this -- and, if so, why?  BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]} \dan{[Don't recall the autocorrelation length. Will?]}
}
\end{figure}

>>>>

\begin{abstract}
Selection among alternative theoretical models given an observed data set is an important challenge in many areas of physics and astronomy.  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an extremely powerful technique for performing Bayesian model selection, but it suffers from a fundamental difficulty: it requires jumps between model parameter spaces, but cannot efficiently explore both parameter spaces at once.  Thus, a naive jump between parameter spaces is unlikely to be accepted in the MCMC algorithm and convergence is correspondingly slow.  Here we demonstrate an interpolation technique that uses samples from single-model MCMCs to propose inter-model jumps from an approximation to the single-model posterior of the target parameter space.  The interpolation technique, based on a kD-tree data structure, is adaptive and efficient in modest dimensionality.  We show that our technique leads to improved convergence over naive jumps in an RJMCMC, and compare it to other proposals in the literature to improve the convergence of RJMCMCs.  We also demonstrate the use of the same interpolation technique as a way to construct efficient ``global'' proposal distributions for single-model MCMCs without prior knowledge of the structure of the posterior distribution, and discuss improvements that permit the method to be used in higher-dimensional spaces efficiently.
\end{abstract}

@@

Here we introduce an alternative technique based on a kD-tree data 
structure to construct an approximation
to each model's posterior parameter distribution.  This improved interpolation method leads to faster convergence of RJMCMC sample chains.
We draw jump proposals into the model from this approximation to its posterior.
Because jumps are proposed preferentially to locations favored by the
single-model posterior, the RJMCMC compares ``good'' locations in
parameter space across all the models, and convergence is generally
rapid.  We have successfully applied this RJMCMC technique to a 10-way
model selection among alternative mass distribution models for
black-hole X-ray binaries \cite{Farr2010}.  We also provide an example
using this method as an ``asymptotically Markovian''
\cite{terBraak2008} jump proposal in the context of a single-model,
nine-dimensional MCMC in Section \ref{sec:higherdim}.

@@

The parameter space of the super-model in an RJMCMC consists of a
discrete parameter that identifies the model, $M_i$, and a set of
continuous parameters appropriate for that model, $\vtheta_i$.  Thus,
each sample consists of a model identifier and a location within the
parameter space of that model, $\{M_i, \vtheta_i\}$.  We perform the
MCMC in the ``super-model'' parameter space just like a regular MCMC;
we propose jumps to different parameters within a model (intramodel
jumps) and jumps to a different model with different parameters
(intermodel jumps).  The acceptance probability for a proposed jump
from $\vtheta^{(j)}_i$ in model $M_i$ to $\vtheta^p_j$ in model $M_j$
becomes
\begin{equation}
p_{\textnormal{accept}} \equiv \min\Bigl(1,
\frac{p(\vtheta^p_j, M_j|d)}{p(\vtheta^{(j)}_i, M_i|d)}
\frac{Q(\vtheta^p_j ,M_j \to
  \vtheta^{(j)}_i, M_i)}{Q(\vtheta^{(j)}_i, M_i \to \vtheta^p_j, M_j)} \Bigr).
\end{equation}
Here the $Q$ factors incorporate both a discrete probability on the
model index, reflecting the probabilistic choice of which model to
jump into, and also a continuous probability density on target model's
parameter space.  For example,
$Q(\vtheta^p_j ,M_j \to \vtheta^{(j)}_i, M_i)$ has a factor for the
probability of proposing a jump to model $i$ when in model $j$, and is
a density on the parameter space of $M_i$.  These densities cancel the
corresponding densities in the ratio of posteriors, making the
acceptance probability a parameterisation-independent scalar.  In the
common special case where the two parameter spaces have equal
dimension and the jump proposal is a diffeomorphism, $\phi$, between
them,
\begin{equation}
  \vtheta^p_j = \phi\left( \vtheta^{(j)}_i \right),
\end{equation}
and therefore 
\begin{equation}
  \vtheta^{(j)}_i = \phi^{-1} \left( \vtheta^p_j \right),
\end{equation}
then the jump proposal ratio reduces to the Jacobian of the
diffeomorphism:
\begin{equation}
  \frac{Q(\vtheta^p_j ,M_j \to
  \vtheta^{(j)}_i, M_i)}{Q(\vtheta^{(j)}_i, M_i \to \vtheta^p_j, M_j)}
= \left| \frac{\partial \phi}{\partial \vtheta^{(j)}_i} \right|.
\end{equation}

 The resulting chain samples from the posterior
$p(M_i, \{\vtheta_i\}|d)$.  As in a usual MCMC, the PDF on the model as a
parameter, with other parameters ignored, is obtained by marginalizing
over the remaining parameters.  The posterior probability of a model
is proportional to the number of counts
%
\be
%
p(M_i|d) = \int d\vtheta_i \frac{L(d|M_i, \vtheta_i) p(\vtheta_i|M_i) p(M_i)}{p(d)}
\approx \frac{N_i}{N},
%
\ee
%
where $N_i$ is the number of RJMCMC samples listing the $i$'th model
and $N$ is the total chain length.  Thus, the probability of a
particular model relative to other models under consideration is given
by the fraction of RJMCMC samples lying in the parameter space of that
model.

@@

In this section, we demonstrate the efficiency of the kD-interpolated
jump proposal on a toy model-comparison problem.  The same algorithm
has been used in real-world settings \cite{Farr2010} and, as discussed
below, is the available in several forms as a software library for
re-use by others.

In the toy model of this section, we draw $N = 100$ simulated data
points from a $N(0,1)$ Gaussian distribution, and then ask whether
these data are better described by a model where they are Gaussian
distributed with unknown mean $\mu$ and standard deviation $\sigma$
%
\be
p(x) = \frac{1}{\sqrt{2\pi} \sigma} \exp\left( - \frac{(x-\mu)^2}{2
    \sigma^2} \right)\ ,
\ee
%
or by a model where they are Cauchy distributed with mode $\alpha$
and width $\beta$
%
\be
p(x) = \frac{1}{\pi \beta \left( 1 + \left(\frac{x - \alpha}{\beta}\right)^2\right)}\ .
\ee
%
We take priors on $\mu$ and $\alpha$ to be uniform in $[-1,1]$, and
priors in $\sigma$ and $\beta$ to be uniform in $[0.5, 1.5]$.  With a
data set of 100 points, the relative uncertainty in determining the
parameters of the underlying distribution is approximately 10\%, so we
expect the posterior probabilities in the $(\mu,\sigma)$ and
$(\alpha,\beta)$ spaces to occupy only a few percent of the prior
volume.  The Cauchy distribution is much broader than the Gaussian (it
has no finite moments), so with equal model priors, the posterior
probability for the Gaussian model over the Cauchy model is extremely
large:
%
\be
\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} \sim 10^9.
\ee
%
In order to ensure that the RJMCMC produces samples in the Cauchy
model at all, we impose a model prior that favors the Cauchy model by
$5 \times 10^8$ relative to the Gaussian.  The evidence ratio between
the models for our chosen data set with these priors is
%
\be \frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)}
\equiv r = 1.15,
\ee
%
yielding a theoretical maximum acceptance rate of inter-model jumps of
$(1+1/r)/2 = 0.93$.

@@

The relative error on the determination of the Bayes factor (evidence ratio) scales
with $1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of inter-model
transitions in the RJMCMC.  Thus, as the acceptance rate of
inter-model jumps goes down, the RJMCMC must run longer to achieve a
desired accuracy in the evidence ratio.  By boosting the acceptance
rate of inter-model jumps, the interpolation method described above
can improve the runtime of an RJMCMC.

@@

\begin{figure}
  \includegraphics[width=0.8\columnwidth]{PCC}
  \caption{\label{fig:PCC} A two-dimensional illustration of the two
    boxes involved in the modified kD proposal.  The larger box
    aligned with the axes is the normal kD-tree box containing the
    given samples, while the tighter box is the ``covariance cell''
    aligned with the eigenvectors of the sample distribution.  The
    modified kd-interpolated jump proposal draws a point from the
    intersection of the two boxes. Tightly-correlated posteriors in
    parameter space such as this are typical of the gravitational-wave
    parameter estimation problem described in \cite{Veitch:2014} and
    in the text.  Without the modification to account for the
    correlated samples, the kD neighbourhood of these points would
    produce a very inefficient jump proposal, since most of the
    bounding-box would contain empty (i.e.\ low-posterior) parameter
    space.}
\end{figure}

@@

Figure \ref{fig:accratio} shows the acceptance ratio for the kD-tree
jump proposal applied as part of a \texttt{LALInferenceMCMC} analysis.
In this example, the kD-tree jump proposal is one of several jump
proposals used during the MCMC, and the kD tree itself is updated with
accepted samples whenever its jump proposal is called. In spite of the
very low acceptance rate of the proposal, applying the kD-tree
proposal to one out of 20 proposed jumps improved the convergence
time---defined as the average time to reach a specified number of
independent samples from the posterior, thinning by the
autocorrelation length as described in \cite{Veitch:2014}---of the
simulation by a factor of two compared to the standard suite of
proposals because it is particularly efficient at producing
``mode-hopping'' jumps, which are difficult to produce with the other
proposals in \texttt{LALInferenceMCMC}.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative acceptance
    ratio for the modified kD-tree jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). The simulation in question was a
    nine-dimensional analysis of a simulated gravitational-wave signal
    injected into synthetic data similar to that taken by the LIGO
    and Virgo detectors \cite{AdvLIGO,AdvVirgo}.  The parameter space
    includes the masses of the compact objects generating the
    gravitational-wave, their location on the sky, distance, orbital
    orientation, the time of signal arrival, and the orbital phase.
    The posterior in this problem has a number of well-separated modes
    in the parameter space which are difficult to jump between using
    traditional jump proposals; in spite of the small acceptance ratio
    of the kD proposal, when applied to one in twenty jumps proposed
    in this simulation, it improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals. 
    The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated.
%\dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).  Is this a fluke, or do you always see this -- and, if so, why?  BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]} \dan{[Don't recall the autocorrelation length. Will?]}
  }
\end{figure}
